---
title: "Rethinking How We Choose Statistical Tests — Solution"
subtitle: "Decision Academy Challenge"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

# How the Data Was Generated

The data was simulated with a known structure so we can compare our statistical conclusions to the ground truth:

- **Control group** (n=50): Normal(50, 10)
- **Treatment group** (n=50): 55% Non-responders from Normal(42, 8) + 45% Responders from Normal(78, 8)
- **True treatment mean**: 0.55 × 42 + 0.45 × 78 = 58.2 (**true effect = +8.2**)
- Site effects: Site A (+0), Site B (−3), Site C (+2)

The treatment genuinely works, but not for everyone — some respond strongly, others see little benefit. This is realistic for many interventions. The question is: which statistical approach recovers this true effect?

```{r simulation-reference, eval=FALSE}
set.seed(394)
N <- 50

# Control
ctrl_scores <- rnorm(N, mean = 50, sd = 10)

# Treatment: bimodal mixture
n_nonresp <- round(N * 0.55)  # 27
n_resp    <- N - n_nonresp     # 23
trt_scores <- c(
  rnorm(n_nonresp, mean = 42, sd = 8),
  rnorm(n_resp, mean = 78, sd = 8)
)
trt_scores <- sample(trt_scores)  # shuffle

cat("True treatment mean:", 0.55 * 42 + 0.45 * 78, "\n")
cat("True difference:", 0.55 * 42 + 0.45 * 78 - 50, "\n")
```

# Part 2: Load the Challenge Data

```{r load-data}
data <- read.csv("challenge_data.csv")
cat("Shape:", nrow(data), "x", ncol(data), "\n")
head(data, 10)
```

```{r descriptives}
ctrl <- data$cognitive_score[data$group == "control"]
trt  <- data$cognitive_score[data$group == "treatment"]

cat("=== Descriptive Statistics ===\n")
cat("Control — mean:", round(mean(ctrl), 2),
    "sd:", round(sd(ctrl), 2),
    "median:", round(median(ctrl), 2), "\n")
cat("Treatment — mean:", round(mean(trt), 2),
    "sd:", round(sd(trt), 2),
    "median:", round(median(trt), 2), "\n")
cat("\nMean difference:", round(mean(trt) - mean(ctrl), 2), "\n")
```

# Part 3: Visualization

```{r histograms, fig.width=12, fig.height=5}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

hist(ctrl, breaks = 15, col = "#0056d2", border = "white",
     main = "Control Group", xlab = "Cognitive Score", xlim = c(20, 100))
abline(v = mean(ctrl), col = "#ab615d", lwd = 2, lty = 2)
legend("topright", legend = paste("Mean =", round(mean(ctrl), 1)),
       col = "#ab615d", lwd = 2, lty = 2, bty = "n")

hist(trt, breaks = 15, col = "#6f8a6d", border = "white",
     main = "Treatment Group", xlab = "Cognitive Score", xlim = c(20, 100))
abline(v = mean(trt), col = "#ab615d", lwd = 2, lty = 2)
abline(v = median(trt), col = "#b08a3a", lwd = 2, lty = 3)
legend("topright",
       legend = c(paste("Mean =", round(mean(trt), 1)),
                  paste("Median =", round(median(trt), 1))),
       col = c("#ab615d", "#b08a3a"), lwd = 2, lty = c(2, 3), bty = "n")

par(mfrow = c(1, 1))
```

The treatment group is clearly **bimodal** — one cluster of non-responders around 40–45 and one cluster of responders around 75–80.

```{r density, fig.width=10, fig.height=5}
plot(density(ctrl), col = "#0056d2", lwd = 2, xlim = c(10, 110),
     main = "Density Comparison: Treatment is Bimodal",
     xlab = "Cognitive Score")
lines(density(trt), col = "#6f8a6d", lwd = 2)
abline(v = mean(ctrl), col = "#0056d2", lty = 2)
abline(v = mean(trt), col = "#6f8a6d", lty = 2)
legend("topright", legend = c("Control", "Treatment"),
       col = c("#0056d2", "#6f8a6d"), lwd = 2, bty = "n")
```

# Part 4: Normality Tests

```{r shapiro}
sw_ctrl <- shapiro.test(ctrl)
sw_trt  <- shapiro.test(trt)

cat("=== Shapiro-Wilk Test ===\n")
cat(sprintf("Control:   W = %.4f, p = %.4f\n", sw_ctrl$statistic, sw_ctrl$p.value))
cat(sprintf("Treatment: W = %.4f, p = %.6f\n", sw_trt$statistic, sw_trt$p.value))
cat("\nBoth groups depart from normality.\n")
cat("A common approach: 'not normal -> use Mann-Whitney U'\n")
cat("Let's see what happens when we try all three tests...\n")
```

# Part 5: The Three Tests

```{r all-tests}
# Student's t-test (equal variances)
t_student <- t.test(trt, ctrl, var.equal = TRUE)

# Welch's t-test (unequal variances)
t_welch <- t.test(trt, ctrl, var.equal = FALSE)

# Mann-Whitney U
mw <- wilcox.test(trt, ctrl, exact = FALSE)

cat("============================================================\n")
cat("TEST RESULTS\n")
cat("============================================================\n")
cat(sprintf("Student's t-test:  t = %.4f, p = %.4f  %s\n",
            t_student$statistic, t_student$p.value,
            ifelse(t_student$p.value < 0.05, "*** SIGNIFICANT", "not significant")))
cat(sprintf("Welch's t-test:    t = %.4f, p = %.4f  %s\n",
            t_welch$statistic, t_welch$p.value,
            ifelse(t_welch$p.value < 0.05, "*** SIGNIFICANT", "not significant")))
cat(sprintf("Mann-Whitney U:    W = %.1f,  p = %.4f  %s\n",
            mw$statistic, mw$p.value,
            ifelse(mw$p.value < 0.05, "*** SIGNIFICANT", "NOT significant")))

cat("\nCOMPARING TO THE TRUE EFFECT (+8.2 points):\n")
cat("Both t-tests detect a significant effect (p ~ 0.03).\n")
cat(sprintf("Mann-Whitney says there is NO effect (p = %.2f).\n", mw$p.value))
cat("\nFollowing 'not normal -> Mann-Whitney' would have missed a real effect.\n")
```

# Part 6: Why Do the Tests Disagree?

**T-tests** compare population **means**. The Central Limit Theorem ensures the sampling distribution of the mean is approximately normal at n=50, regardless of the data shape.

**Mann-Whitney U** tests **stochastic dominance**: P(X > Y) ≠ 0.5. With bimodal data, 55% of treatment observations sit *below* the control center, weakening the stochastic signal even though the overall mean is higher.

```{r bootstrap, fig.width=10, fig.height=5}
set.seed(0)
n_boot <- 10000
boot_diffs <- replicate(n_boot, {
  bc <- sample(ctrl, replace = TRUE)
  bt <- sample(trt, replace = TRUE)
  mean(bt) - mean(bc)
})

ci <- quantile(boot_diffs, c(0.025, 0.975))

hist(boot_diffs, breaks = 50, col = "#b08a3a", border = "white",
     main = "Bootstrap Distribution of Mean Difference",
     xlab = "Treatment - Control (mean)")
abline(v = 0, col = "#ab615d", lwd = 2)
abline(v = mean(boot_diffs), col = "#1f1f1c", lwd = 2, lty = 2)

cat(sprintf("Bootstrap 95%% CI: [%.2f, %.2f]\n", ci[1], ci[2]))
cat("The CI does not contain 0 — confirming the treatment effect.\n")
```

# Part 7 (Level 2 (Advanced)): ANCOVA

```{r ancova-baseline}
model1 <- lm(cognitive_score ~ group + baseline_score, data = data)
cat("=== ANCOVA: group + baseline_score ===\n")
summary(model1)
```

```{r ancova-full}
model2 <- lm(cognitive_score ~ group + baseline_score + site, data = data)
cat("=== ANCOVA: group + baseline_score + site ===\n")
summary(model2)

cat("\nKEY INSIGHT:\n")
cat("Adding site increases the treatment effect from ~6.3 to ~8.0\n")
cat("and makes it more significant (p = 0.007 vs 0.033).\n")
cat("This is very close to the TRUE effect of +8.2.\n")
cat("Site was a source of unexplained variance masking part of the effect.\n")
```

# Comparison to the True Effect

| Approach | Estimated effect | p-value | True effect = 8.2 |
|----------|----------------|---------|------------|
| Mann-Whitney U | (not a mean difference) | 0.398 | Misses it entirely |
| Welch's t-test | +6.3 | 0.033 | Detects, underestimates |
| ANCOVA (+ baseline) | +6.3 | 0.028 | Detects, underestimates |
| ANCOVA (+ baseline + site) | +8.0 | 0.007 | Closest to truth |

The ANCOVA with site effects recovers +8.0 — remarkably close to the true +8.2 — with the strongest evidence (p = 0.007).

**Key Takeaways:**

1. "Not normal" does not mean "use Mann-Whitney." Match the test to your question.
2. Mann-Whitney tests stochastic dominance, not means.
3. T-tests are robust to non-normality at n = 50 (CLT).
4. Visualize first. The bimodality was obvious in histograms.
5. Design beats test selection. ANCOVA recovered the true effect almost exactly — far better than any simple test.
